Outline of Ideas for BTC Transaction Frequency

### 2015-10-05 outline/outline.tex
Added some detail about simulations

Crucial constraints on the models we can fit come from the data obtained from the block chain. Since Bitcoin is a distributed network without a home node, there is no official timestamp for when each block arrives on the chain. Instead, the block miner's local time is recorded. Two constraints on the reported time that the block was mined are built into the Bitcoin protocol. First, any block submitted to the chain with a local time less than the median local times of the last 11 blocks is rejected. Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node, with a maximum offset of 70 minutes. Because of the nature of the blocks' timestamps, occasionally a block has a timestamp which is earlier than one or more previous blocks in the chain. Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid timestamps define lower and upper bounds for the measurement error distribution. Unfortunately network adjusted time is not recorded in the blockchain, so the upper bound is not observed, but the upper bound is at our disposal.

The second major constraint that the block chain imposes on us is that transaction data is only measured once a block is completed. So at random measurement times which we only observe with error, we observe the amount of transaction data since the previous measurement time --- though we observe this without error. These two constraints motivate a class of models for modeling Bitcoin transaction data. Let $\tau_k$ denote the actual time of the arriaval of the $k$'th block, $\delta_k$ the amount of time it is the current block so that $\tau_{k+1}=\tau_k + \delta_k$, and $t_k$ the observed time in the blockchain. Further, let $x_\tau$ denote the amount data in the chain at time $\tau$ so that $x_{\tau_{k+1}} - x_{\tau_k}$ denotes the amount of data in block $k$. Now suppose the amount of transaction data in MBs is distributed according to some stochastic process $dx_\tau$ so that $x_{\tau + \delta} - x_\tau = \int_\tau^{\tau + \delta}dx_\tau$, a stochastic integral. Then the amount of transaction data that arrives in $k$'th block is given by $\int_{\tau_k}^{\tau_{k}+\delta_k}dx_\tau$.

Most of the interesting modeling choices come from modeling $dx_\tau$. Conventional Brownian motion is inappropriate since the increments should be nonnegative. On the other hand, geometric Brownian motion is a poor choice since it cannot start from zero. Instead, we consider the classes of non-negative infinite activity L\'{e}vy processes with closed form solutions for $x_{\tau}$. The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continous time and has independent and stationary increments with $x_0=0$.  For example, the Gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$. We use the shape-rate parameterization of the gamma distribution so that if $x\sim G(alpha,\beta)$, it has the density
\[
p(x) = \frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-x\beta}
\]
for $x>0$ with mean $\alpha/\beta$ and variance $\alpha/\beta^2$.

While all L\'{e}vy processes are infinitely divisible, the gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form. Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$. The inverse Gaussian process is another L\'{e}vy process which has a closed form solution for the distribution of $x_\tau$. The inverse Gaussian density is
\[
p(x) = \left[\frac{\lambda}{2\pi x^3}\right]^{1/2}\exp\left[\frac{-\lambda(x - \mu)^2}{2\mu^2x}\right]
\]
for $x>0$ with mean parameter $\mu>0$, shape parameter $\lambda>0$, and variance $\mu^3/\lambda$. Then increments of the inverse Gaussian process are inverse Gaussian distributed, viz. $x_{\tau + \delta} - x_\tau \sim IG(\mu\delta,\lambda\delta^2)$.

It is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form since the observation times are observed with error since this simplifies Markov chain Monte Carlo methods that may be used to fit the model. The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g. seasonal variation or stochastic volatility, but most reasonable models on these parameters will not allow for closed form distributions of the parameter increments.

On the data level we have measurement error with a lower bound, motivating a class of models of the form 
\[
t_{k+1} |t_{1:k}, \tau_{1:K} \stackrel{ind}{\sim} p(t|\tau_{k+1}, \phi)\mathbbm{1}\left\{t > m_{k+1})\right\}
\]
where $p(t|\tau,\phi)$ is a class of probability densities on $t>0$ with centrality parameter $\tau$ and dispersion parameter $\phi$, and $m_k = \mathrm{median}(t_{k - 0:10})$. For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision. This interpretation of these parameter applies for the {\it untruncated} dsitribution $p(t|\tau,\phi)$, but not necessarily for the condtional dsitribution of $t_k$.

For example, consider the following parameterized modeling using the structure discussed above. For block $k=1,2,\dots,K$
\begin{align*}
t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\alpha\delta_k,\beta)\\
\delta_{k-1} = \tau_k - \tau_{k-1} &\stackrel{iid}{\sim} Exp(1/10)
\end{align*}
where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$. Note that $x_{\tau_{k+1}} - x_{\tau_k}$ depends on $\delta_k$ because the $k$'th block arrives at $\tau_k$ and departs at $\tau_{k+1}=\tau_k + \delta_k$ --- the block has $\delta_k$ minutes to accumulate transaction data.

\section{Put this where it needs to go}

The quantity we are interested in is thus
\[
P\left(\int_{\tau_k}^{\tau_{k}+\delta_k}dx_\tau > 1\right)
\]
for some choice of $k$.

For some distributional forms for the process $dx_\tau$ it may be possible to compute these quantities analytically, but in general it may be difficult. Using a Monte Carlo simulation of $N$ draws, we can approximate these probabilities by
\[
\frac{1}{N}\sum_{i=1}^N\mathbbm{1}\left\{\int_{\tau_k^{(i)}}^{\tau_{k}^{(i)}+\delta_k^{(i)}}dx_\tau^{(i)} > 1\right\}
\]
where $i$ indexes the Monte Carlo simulations of each of the random quantities inside the indicator function. This formula assumes that we can perfectly simulate the continous sample paths of $dx_\tau$, which may require further approximation.

>>>>

Crucial constraints on the models we can fit come from the data obtained from the block chain. Since Bitcoin is a distributed network without a home node, there is no official timestamp for when each block arrives on the chain. Instead, the block miner's local time is recorded. Two constraints on the reported time that the block was mined are built into the Bitcoin protocol. First, any block submitted to the chain with a local time less than the median local times of the last 11 blocks is rejected. Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node, with a maximum offset of 70 minutes. Because of the nature of the blocks' timestamps, occasionally a block has a timestamp which is earlier than one or more previous blocks in the chain. Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid timestamps define lower and upper bounds for the measurement error distribution. Unfortunately network adjusted time is not recorded in the blockchain, so the upper bound is not observed, but the lower bound is at our disposal.

The second major constraint that the block chain imposes on us is that transaction data is only measured once a block is completed. So at random measurement times which we only observe with error, we observe the amount of transaction data since the previous measurement time --- though we observe this without error. These two constraints motivate a class of models for modeling Bitcoin transaction data. Let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time it is the current block so that $\tau_{k+1}=\tau_k + \delta_k$, and $t_k$ the observed arrival time in the blockchain. Further, let $x_\tau$ denote the amount data in the chain at time $\tau$ so that $x_{\tau_{k+1}} - x_{\tau_k}$ denotes the amount of data in block $k$. Now suppose the amount of transaction data in MBs is distributed according to some stochastic process $dx_\tau$ so that $x_{\tau + \delta} - x_\tau = \int_\tau^{\tau + \delta}dx_\tau$, a stochastic integral. Then the amount of transaction data that arrives in $k$'th block is given by $\int_{\tau_k}^{\tau_{k}+\delta_k}dx_\tau$.

Most of the interesting modeling choices come from modeling $dx_\tau$. Conventional Brownian motion is inappropriate since the increments should be nonnegative. On the other hand, geometric Brownian motion is a poor choice since it cannot start from zero. Instead, we consider the classes of non-negative infinite activity L\'{e}vy processes with closed form solutions for $x_{\tau}$. The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continous time and has independent and stationary increments with $x_0=0$. Brownian motion is the most common example. A better example for our purposes is the Gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$. We use the shape-rate parameterization of the gamma distribution so that if $x\sim G(alpha,\beta)$, it has the density
\[
p(x) = \frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-x\beta}
\]
for $x>0$ with mean $\alpha/\beta$ and variance $\alpha/\beta^2$.

While all L\'{e}vy processes are infinitely divisible, the gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x\sim G(\alpha_x,\beta)$ independent of $y\sim G(alpha_y,\beta)$, then $x+y \sim G(\alpha_x + \alpha_y, \beta)$. Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$. The inverse Gaussian process is another L\'{e}vy process which has a closed form solution for the distribution of $x_\tau$. The inverse Gaussian density is
\[
p(x) = \left[\frac{\lambda}{2\pi x^3}\right]^{1/2}\exp\left[\frac{-\lambda(x - \mu)^2}{2\mu^2x}\right]
\]
for $x>0$ with mean parameter $\mu>0$, shape parameter $\lambda>0$, and variance $\mu^3/\lambda$. Then increments of the inverse Gaussian process are inverse Gaussian distributed, viz. $x_{\tau + \delta} - x_\tau \sim IG(\mu\delta,\lambda\delta^2)$.

It is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form since the observation times are observed with error. This simplifies Markov chain Monte Carlo methods that may be used to fit the model by allowing us to exactly simulate $x_{\tau_k + \delta_k}$ without resorting to approximation. The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g. seasonal variation or stochastic volatility, but many reasonable models on these parameters will not allow for closed form distributions of the parameter increments.

On the data level we have measurement error with a lower bound, motivating a class of models of the form 
\[
t_{k+1} |t_{1:k}, \tau_{1:K} \stackrel{ind}{\sim} p(t|\tau_{k+1}, \phi)\mathbbm{1}\left\{t > m_{k+1})\right\}
\]
where $p(t|\tau,\phi)$ is a class of probability densities on $t>0$ with centrality parameter $\tau$ and dispersion parameter $\phi$, and $m_k = \mathrm{median}(t_{k - 0:10})$. For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision. This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the condtional dsitribution of $t_k$. A natural choice here is another gamma distribution, since the untruncated distribution still must be non-negative. However the gamma causes computational problems since it forces the latent measurement times, $\tau_k$ to enter the gamma function. A similar model that avoids this problem is the lognormal distribution.

To start, consider the following parameterized model using the structure discussed above. For block $k=1,2,\dots,K$
\begin{align*}
t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\beta\mu\delta_k,\beta)\\
\delta_{k-1} = \tau_k - \tau_{k-1} &\stackrel{iid}{\sim} Exp(1/10)
\end{align*}
where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$. Note that $x_{\tau_{k+1}} - x_{\tau_k}$ depends on $\delta_k$ because the $k$'th block arrives at $\tau_k$ and departs at $\tau_{k+1}=\tau_k + \delta_k$ --- the block has $\delta_k$ minutes to accumulate transaction data.

\section{The key question}
As Bitcoin becomes more popular, the transaction data rate increases. Since existing Bitcoin rules limit the number of transactions per block to one MB, at some point the blocks will begin filling up. The purpose of modeling Bitcoin's transaction data rate is to estimate the probability of seeing a full block under a variety of assumptions. Starting from the base model above, the quantity we are interested in is thus
\[
P\left(\int_{\tau_k}^{\tau_{k}+\delta_k}dx_\tau > 1\right)
\]
for some choice of $k$.

For some distributional forms for the process $dx_\tau$ it may be possible to compute these quantities analytically, but in general it may be difficult. Using a Monte Carlo simulation of $N$ draws, we can approximate these probabilities by
\[
\frac{1}{N}\sum_{i=1}^N\mathbbm{1}\left\{\int_{\tau_k^{(i)}}^{\tau_{k}^{(i)}+\delta_k^{(i)}}dx_\tau^{(i)} > 1\right\}
\]
where $i$ indexes the Monte Carlo simulations of each of the random quantities inside the indicator function. The key parameter we will control in these simulations is the mean transaction data rate, $\mu$, or the distribution of this rate. Starting with the basic model fit to, say, the last 1000 blocks on the blockchain (approximately one week of elapsed time), we can control this parameter in several ways.

The simplest way to compute parameter estimates based on fitting model, e.g. maximum likelihood estimates, posterior means, or posterior medians. Then estimate the probability of a block filling up by simulating the model conditional on these parameter values. In order to take into account increased Bitcoin usage, we can manually increase the mean transaction data rate parameter, $\mu$, while holding other parameters constant.

This approach has two major shortfalls: 1) it ignores parameter uncertainty, and 2) it assumes that the other parameters will remain constant as $\mu$ increases. We can mitigate 1) while making 2) if not worse, different. Specifically, suppose we fit the model and have access to the full (approximate) posterior distribution of the model parameters. Then we can simulate from the posterior of the parameters and for each simulated parameter, simulate the model in order to estimate the probability of a block filling up. This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\mu$'s distribution were different, e.g. its mean $\bar{\mu}$. We estimate the probability of a block filling using both approaches, but because of the required assumptions the estimates become less credible for values of $\mu$ or $\bar{\mu}$ further away from the estimated values.
